Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-02-02 2.background/2.Gaussian-distributions.tex
working on ch. 6
implementing my corrections.

{\bf Slepian's lemma and the Sudakov-Fernique inequality.} 
The following two result will play fundamental roles in our characterization of uniform relative stability for Gaussian triangular arrays in 
Chapter \ref{chap:URS}. 
% Many more details and deeper insights can be found for example in the monographs 
%\cite{leadbetter2012extremes} and \cite{adler2009random}, among others. 
The first is the celebrated result due to \citet{slepian1962one}.

\begin{theorem}[Slepian's Lemma]\label{thm:Slepian-lemma} Let $\epsilon = (\epsilon(i))_{i=1}^p$ and $\eta = (\eta(i))_{i=1}^p$ be 
two multivariate normally distributed random vectors with zero means $\E [\epsilon(i)] = \E[\eta(i)] = 0$.

@@

This result implies in particular that $M_\epsilon:= \max_{i=1,\cdots,p}\epsilon(i)$ dominates stochastically $M_\eta:= \max_{i=1,\cdots,p} \eta(i)$.
Thus, for example, the maximum of iid Gaussians are stochastically larger the maxima any positively correlated Gaussian vector 
with the same marginal distributions.

@@

The following result, known as the Sudakov-Fernique inequality is similar in spirit to Slepian's lemma but does not assume that the Gaussian 
vectors are centered. For a proof and, in fact, a more general result, see e.g., Theorem 2.2.5 on page 61 in \cite{adler2009random}.

@@

If for all $i,j=1,\cdots,p$, we have 
$$
\E [ \epsilon(i) ] = \E [\eta(i)] \ \ \mbox{ and }\ \ \E [ (\epsilon(i)-\epsilon(j))^2] \le \E [ (\eta(i)-\eta(j))^2],
$$
then for $M_\epsilon = \max_{i=1,\cdots,p}\epsilon(i)$ and $M_\eta = \max_{i=1,\cdots,p} \eta(i)$, we have 
$$
 \E [ M_\epsilon ] \le \E [M_\eta].
$$
\end{theorem}

>>>>

{\bf Slepian's lemma and the Sudakov-Fernique inequality.} 
The following two results will be instrumental in our characterization of uniform relative stability for 
Gaussian triangular arrays in Chapter \ref{chap:URS}.  
% Many more details and deeper insights can be found for example in the monographs 
%\cite{leadbetter2012extremes} and \cite{adler2009random}, among others. 
The first is the celebrated result due to \citet{slepian1962one}.

\begin{theorem}[Slepian's lemma]\label{thm:Slepian-lemma} Let $\epsilon = (\epsilon(i))_{i=1}^p$ and $\eta = (\eta(i))_{i=1}^p$ be 
two multivariate normally distributed random vectors with zero means $\E [\epsilon(i)] = \E[\eta(i)] = 0$.

@@

This result implies in particular that $M_\epsilon:= \max_{i=1,\cdots,p}\epsilon(i)$ dominates stochastically $M_\eta:= \max_{i=1,\cdots,p} \eta(i)$ in the
sense that
\begin{equation}\label{e:Slepian-lemma-maxima}
 \P[ M_\eta > u] \le \P [M_\epsilon > u],\ \ \mbox{ for all }u\in\mathbb R.
\end{equation}
In this case, we shall write $M_\eta \stackrel{d}{\le} M_\epsilon$.  This result shows, for example, that the maximum of iid Gaussians is 
stochastically larger the maximum of any positively correlated Gaussian vector with the same marginal distributions.

@@

The following result, known as the Sudakov-Fernique inequality, is similar in spirit to Slepian's lemma but it does not assume that the Gaussian 
vectors are centered and yields a weaker conclusion -- an inequality between expectations. For a proof, many insights, and, in fact, a more 
general result, see e.g., Theorem 2.2.5 on page 61 in \cite{adler2009random}.

@@

If for all $i,j=1,\cdots,p$, we have 
$$
\E [ \epsilon(i) ] = \E [\eta(i)] \ \ \mbox{ and }\ \ \E [ (\eta(i)-\eta(j))^2] \le \E [ (\epsilon(i)-\epsilon(j))^2],
$$
then for $M_\epsilon = \max_{i=1,\cdots,p}\epsilon(i)$ and $M_\eta = \max_{i=1,\cdots,p} \eta(i)$, we have 
$$
 \E [ M_\eta ] \le \E [M_\epsilon].
$$
\end{theorem}

### 2021-02-02 5.URS/URS.tex
working on ch. 6
implementing my corrections.

In this chapter, we establish a complete characterization of URS for Gaussian arrays in terms of a simple condition on the covariance structures.
The condition is as follows.

@@

\begin{definition}[UDD\,$^\prime$]\label{d:UDD-prime}
For $\epsilon_p \sim \mathrm{N}(0,\Sigma_p)$ with unit variances, an array ${\cal E} = \left(\epsilon_p(i)\right)_{i=1}^p$ is said to satisfy the UDD\,$^\prime$ condition if there 
exist:
\begin{enumerate}
    \item[(i)] permutations $l = l_p$ of $\{1,\ldots,p\}$, for all $p\in\N$, and
    \item[(ii)] a non-negative sequence $(r_n)_{n=1}^\infty$ converging to zero $r_n\to 0$, as $n\to\infty$,
\end{enumerate}
such that 
\begin{equation} \label{eq:weak-correlation}
    \sup_{p\in\N} |\Sigma_p\left(i',j'\right)| \le r_{|i-j|}.
\end{equation}
where $i' = l(i)$, $j' = l(j)$, for all $i,j\in\{1,\ldots,p\}$.
\end{definition}

@@

We now state the main result of this section: a Gaussian sequence is URS if and only if it is UDD.
The URS condition essentially requires that the dependencies decay in a uniform fashion, the rate at which dependence decay does \emph{not} matter.

@@

Returning again to the study of support recovery problems, Theorem \ref{thm:Gaussian-weak-dependence} and the necessary condition for exact support recovery in Theorem \ref{thm:necessary} yields the following result.

@@

As a counterpart to Remark \ref{rmk:dependence-assumptions}, we demonstrate the tightness of the dependence conditions in Theorem \ref{thm:necessary}.
Specifically, we demonstrate that if the URS dependence condition is violated, then it may be possible to recover the support of weaker signals below the boundary.

\begin{example} \label{exmp:counter-example}
Suppose ${\cal E} = \left(\epsilon_p(i)\right)_{i=1}^p$ is Gaussian, and is comprised of $\lfloor p^{1-\beta}\rfloor$ blocks, each of size at least $\lfloor p^\beta \rfloor$; 
let the elements of each block have correlation 1, and let elements from different blocks be independent. 
If $\underline{r} \ge 4(1-\beta)$, then the procedure $\widehat{S} = \big\{i:x(i)>\sqrt{2(1-\beta)\log{p}}\big\}$ yields $\mathbb P[\widehat{S} = S] \to 1$. 
This requirement on signal size is strictly weaker than that of the strong classification boundary, since $4(1-\beta) < (1 + \sqrt{1-\beta})^2$ on $\beta\in(0,1)$.
\end{example}

The above example shows that if the correlations of the Gaussian errors do not decay in a uniform fashion (UDD fails), then we can do substantially better in terms of support recovery.
The claims in the example are verified in Section \ref{suppsec:proofs}, while numerical simulations of this example can be found in Section \ref{suppsec:numerical}.

@@

We conclude with a brief discussion on the relationships between UDD and other dependence conditions in the context of extreme value theory, before proceeding to the proof of Theorem \ref{thm:Gaussian-weak-dependence}.

>>>>

Recall that the notion of uniform relative stability (URS) in Definition \ref{def:URS} is the key to the necessary conditions for exact support recovery 
established in Theorem \ref{thm:necessary}.
In this chapter, we provide a complete characterization of the class of URS Gaussian arrays in terms of a simple condition on their covariance structure.
The condition is as follows.

@@

\begin{definition}[UDD\,$^\prime$]\label{d:UDD-prime}
For $\epsilon_p \sim \mathrm{N}(0,\Sigma_p)$ with unit variances, an array ${\cal E} = \left(\epsilon_p(i)\right)_{i=1}^p$ is said to satisfy the UDD\,$^\prime$ condition if there 
exist:
\begin{enumerate}
    \item[(i)] permutations $l_p$ of $\{1,\ldots,p\}$, for all $p\in\N$, and
    \item[(ii)] a non-negative sequence $(r_n)_{n=1}^\infty$ converging to zero $r_n\to 0$, as $n\to\infty$,
\end{enumerate}
such that 
\begin{equation} \label{eq:weak-correlation}
    \sup_{p\in\N} |\Sigma_p\left(i',j'\right)| \le r_{|i-j|}.
\end{equation}
where $i' = l_p(i)$, $j' = l_p(j)$, for all $i,j\in\{1,\ldots,p\}$.
\end{definition}

@@

We now state the main result of this chapter.  It states that a Gaussian array is URS if and only if it is UDD.
The URS condition essentially requires that the dependencies decay in a uniform fashion, the rate at which dependence decay 
does \emph{not} matter.

@@

Returning again to the study of support recovery problems, Theorem \ref{thm:Gaussian-weak-dependence} and the necessary condition for exact support recovery in Theorem \ref{thm:necessary} yield the following result.

@@

One may ask, whether the UDD (equivalently, URS) condition can be relaxed further for the phase-transition result in Theorem \ref{thm:necessary} to 
hold.  As a counterpart to Remark \ref{rmk:dependence-assumptions}, we demonstrate next that the dependence conditions in Theorem \ref{thm:necessary} 
are nearly optimal.  Specifically, we show that if the URS dependence condition is violated, then it may be possible to recover the support 
of weaker signals, falling below the boundary.   The main idea is to use the equivalence of URS and UDD to construct a Gaussian error array,
whose correlations do not decay in a uniform fashion (UDD fails). As we will see, in such a case one can do substantially better in terms of 
support recovery.  This shows that the URS condition is nearly optimal in the Gaussian setting.  
Numerical simulations illustrating this example can be found in Section \ref{suppsec:numerical}, below.

\begin{example}[On the tightness of the URS condition for exact support recovery] \label{exmp:counter-example}
Suppose ${\cal E} = \left(\epsilon_p(i)\right)_{i=1}^p$ is Gaussian, and is comprised of $\lfloor p^{1-\beta}\rfloor$ blocks, each of size at least $\lfloor p^\beta \rfloor$.  Let the elements within each block have correlation 1, and let the elements from different blocks be independent. 
If $\underline{r} \ge 4(1-\beta)$, then the procedure 
$$
 \widehat{S} = \big\{i:x(i)>\sqrt{2(1-\beta)\log{p}}\big\}
 $$ 
yields exact support recovery, i.e., $\mathbb P[\widehat{S} = S] \to 1$, as $p\to\infty$.  This requirement on the signal size is strictly 
weaker than that of the strong classification boundary, since $4(1-\beta) < (1 + \sqrt{1-\beta})^2$ on $\beta\in(0,1)$.
\end{example} 
\begin{proof}[Example \ref{exmp:counter-example}]
Let $t_p^* = \sqrt{2(1-\beta)\log{p}}$  and observe that $\widehat{S} = \{j:x(j)>t_p^*\}$.
Analogous to \eqref{eq:Bonferroni-FWER-control} in the proof of Theorem \ref{thm:sufficient}, we have
\begin{align*}
    \P\left[\widehat{S} \subseteq S\right] 
        &= 1 - \P\left[\max_{j\in S^c}x(j) > t_p^*\right] 
        = 1 - \P\left[\max_{j\in S^c}\epsilon(j) > t_p^*\right] \nonumber \\
      % \ge 1 - \P\left[\max_{j\in\{1,\ldots,p\}}\epsilon(j) > t_p\right] \nonumber \\
        &\ge 1 - \P\left[\max_{j\in\{1,\ldots,p\}}\epsilon(j) > t_p^*\right] 
        \ge 1 - \P\left[\max_{j\in\{1,\ldots,\lfloor p^{1-\beta}\rfloor\}}\widetilde{\epsilon}(j) > t_p^*\right]
\end{align*}
where $\left(\widetilde{\epsilon}\right)_{j=1}^{\lfloor p^{1-\beta}\rfloor}$'s are independent Gaussian errors; in the last inequality we used the assumption that there are at most $\lfloor p^{1-\beta}\rfloor$ independently distributed Gaussian errors in $\left(\epsilon_p(j)\right)_{j=1}^p$.
By Example \ref{exmp:FWER-controlling_procedures} (with $\lfloor p^{1-\beta}\rfloor$ taking the role of $p$), we know that the FWER goes to 0 at a rate of 
$\left(2\log{\lfloor p^{1-\beta}\rfloor}\right)^{-1/2}$.
Therefore, the probability of no false inclusion converges to 1.

On the other hand, since the signal sizes are no smaller than $(\nu\underline{r}\log p)^{1/\nu} = \sqrt{2\underline{r}\log p}$ (for $\nu = 2$), similar to 
\eqref{eq:sufficient-proof-eq1}, we obtain
\begin{align}
    \P\left[\widehat{S} \supseteq S\right] 
    &\ge \P\left[\min_{j\in S}\epsilon(j) > \sqrt{2(1-\beta)\log{p}} - \sqrt{2\underline{r}\log{p}} \right] \nonumber \\
    &= \P\left[\max_{j\in S}\left(-\epsilon(j)\right) < \sqrt{2\log{p}}\left(\sqrt{\underline{r}}-\sqrt{1-\beta}\right) \right] \nonumber \\
    &= \P\left[\frac{\max_{j\in S}(-\epsilon(j))}{u_{|S|}} < \frac{\sqrt{\underline{r}}-\sqrt{1-\beta}}{\sqrt{1-\beta}}\left(1+o(1)\right) \right], \label{eq:sufficient-proof-counter-example}
\end{align}
where in the last line we used the quantiles \eqref{eq:AGG-quantiles}.
Since the minimum signal size is bounded below by $\underline{r} > 4(1-\beta)$, the right-hand-side of the inequality in \eqref{eq:sufficient-proof-counter-example} converges to a constant strictly larger than 1. While the left-hand-side, by Slepian's lemma (recall 
Theorem \ref{thm:Slepian-lemma} and Relation \ref{e:Slepian-lemma-maxima}), is stochastically smaller than a r.v.\ going to 1.  Namely, we have
\begin{equation}
  \frac{1}{u_{|S|}} \max_{j\in S}(-\epsilon(j)) \stackrel{d}{\le} \frac{1}{u_{|S|}} \max_{j\in S} \epsilon^*(j) \stackrel{\P}{\longrightarrow} 1,
\end{equation}
where $\left({\epsilon^*}\right)_{j=1}^{\lfloor p^{1-\beta}\rfloor}$'s are independent Gaussian errors.
Therefore the probability in \eqref{eq:sufficient-proof-counter-example} must also converge to 1.
\end{proof}

@@

Before proceeding to the proof of Theorem \ref{thm:Gaussian-weak-dependence}, we will briefly discuss 
the relationships between UDD and other dependence conditions in the context of extreme value theory. The main idea
we would like to convey is that UDD (and equivalently URS) is an exceptionally mild condition on the dependence of the array.
